# Probability Rules

## Multiplication Rule

In this section we again use the blood type data in Table 5.1 below.

```{r, echo=FALSE}
table <- data.frame(a = c("A", "B", "AB", "O"), b = c(360, 76, 25, 390), d = c(60, 14, 5, 70))
table %>% kbl(caption = "Frequencies by ABO and Rh Groups", col.names = c("Group", "$+$", "$-$")) %>% kable_styling(full_width = FALSE)
```

In the previous section we computed the conditional probability

$$
\mbox{P(A | }-) = \frac{60}{149} = 0.403
$$

based on the data in Table 1.  
The numerator 60 came from the number of Rh $-$ people who are also ABO group A, 
and the denominator 149 is the total number of Rh $-$ people. 
If we divide the numerator and denominator by 1000, then the quotient is not changed
so that we have

$$
\mbox{P(A | }-) = \frac{60/1000}{149/1000} 
= \frac{0.060}{0.149} = \frac{\mbox{P(A and }-)}{\mbox{P(}-)}
$$

This illustrates a general property of probability: If X and Y represent possible 
outcomes from a random event, then the probability of X given Y is

$$
\mbox{P(X | Y)} = \frac{\mbox{P(X and Y)}}{\mbox{P(Y)}}
$$

Multiplying on both sides of the above equation by P(Y) gives

$$
\mbox{P(X and Y)} = \mbox{P(X | Y)}\mbox{P(Y)}  
$$

This is called the **multiplication rule**, and it allows us to move back and forth
between joint probabilities and conditional probabilities.  

Let's try this out, using Table 5.2 (which we saw previously).  

```{r, echo=FALSE}
table <- data.frame(a = c("A", "B", "AB", "O"), b = c(0.360, 0.076, 0.025, 0.390), d = c(0.060, 0.014, 0.005, 0.070))
table %>% kbl(caption = "Probabilities by ABO and Rh Groups", col.names = c("Group", "$+$", "$-$")) %>% kable_styling(full_width = FALSE)
```

---

## Exercises

1. $\text{Put exercises here}$
