# Probability Rules

## Multiplication Rule

In this section we again use the blood type data in Table 5.1 below.

```{r, echo=FALSE}
table <- data.frame(a = c("A", "B", "AB", "O"), b = c(360, 76, 25, 390), d = c(60, 14, 5, 70))
table %>% kbl(caption = "Frequencies by ABO and Rh Groups", col.names = c("Group", "$+$", "$-$")) %>% kable_styling(full_width = FALSE)
```

In the previous section we computed the conditional probability

$$
\mbox{P(A | }-) = \frac{60}{149} = 0.403
$$

based on the data in Table 1.  The numerator 60 came from the number of Rh $-$ 
people who are also ABO group A, 
and the denominator 149 is the total number of Rh $-$ people. 
If we divide the numerator and denominator by 1000, then the quotient is not changed
so that we have

$$
\mbox{P(A | }-) = \frac{60/1000}{149/1000} 
= \frac{0.060}{0.149} = \frac{\mbox{P(A and }-)}{\mbox{P(}-)}
$$

This illustrates a general property of probability: If X and Y represent possible 
outcomes from a random event, then the probability of X given Y is

$$
\mbox{P(X | Y)} = \frac{\mbox{P(X and Y)}}{\mbox{P(Y)}}
$$

Multiplying on both sides of the above equation by P(Y) gives

$$
\mbox{P(X and Y)} = \mbox{P(X | Y)}\mbox{P(Y)}  
$$

This is called the **multiplication rule**, and it allows us to move back and forth
between joint probabilities and conditional probabilities.  

Let's try the multiplicatioon rule out, using the probabilities in 
Table 5.2 (which we saw previously).  

```{r, echo=FALSE}
table <- data.frame(a = c("A", "B", "AB", "O"), b = c(0.360, 0.076, 0.025, 0.390), d = c(0.060, 0.014, 0.005, 0.070))
table %>% kbl(caption = "Probabilities by ABO and Rh Groups", col.names = c("Group", "$+$", "$-$")) %>% kable_styling(full_width = FALSE)
```

Let's start by finding $\mbox{P(B}|+)$.  Apply the above formula, we have

$$
\mbox{P(B}|+) = \frac{\mbox{P(B and }+)}{\mbox{P}(+)} = \frac{0.076}{0.36+0.076+0.025+0.39}
= \frac{0.076}{0.851} = 0.0893
$$

In a similar manner, we compute $\mbox{P(B}|-)$ with

$$
\mbox{P(B}|-) = \frac{\mbox{P(B and }-)}{\mbox{P}(-)} = \frac{0.014}{0.06+0.014+0.005+0.07}
= \frac{0.014}{0.149} = 0.0940
$$

We see that $\mbox{P(B}|+)$ and $\mbox{P(B}|-)$ are not the same.  Thus
the probability of being ABO blood group B depends on whether one is Rh group
$+$ or $-$.

### Sample Questions

<style>
hide {
  background-color: #d6d6d6;
  color: #d6d6d6;
}
hide:hover {
  background-color: white;
  color: black;
}
</style>

**1.** Use the probabilities in Table 5.2 to compute $\mbox{P(O | }+)$

**Answer:** <hide>From Table 1 we see that there are a total of
$360 + 76 + 25 + 390 = 851$ in the Rh $+$ group.  Among these are
390 that are ABO group O.  Therefore 
$\mbox{P(O | }+) = \frac{390}{851} = 0.458$.<hide>

**2.** Update

**Answer:** <hide>Update..<hide>

---

## Independence

Section on independence here

## Exercises

1. $\text{Put exercises here}$
