# Graphing Distributions: Qualitative Variables

 <font size="2"> Note: Portions below draw on content from
 *Online Statistics Education: A Multimedia Course of Study* 
 (http://onlinestatbook.com/) Project Leader: David M. Lane, Rice University </font>

## Introduction

Suppose that we have a community of 500 people.  Each is classified based on 
their ABO blood group, which is one of A, B, AB, or O.  Below we consider
graphical methods for displaying the results of the blood group classifications.
This starts with tables, and then continues on to how to graph data that 
fall into a small number of categories. 

This is an example of **qualitative data** (also known as **categorical data**).  
One characteristic of such data is
that the different values do not come with any pre-established ordering.
This can be contrasted with quantitative data, such as the weight of a bag of an 
unknown substance, which does have a natural ordering with respect to 
different weights. (Another way to think about the distinction: it makes sense
to compute the average of quantitative data but usually it does not for
qualitative data.)

## Frequency Tables

All of the graphical methods shown in this section are derived from frequency 
tables. Table 1 shows a frequency table for the results of the ABO blood group
classification.  It also shows the 
relative frequencies, which are the proportion classified in each category. 
For example, the relative frequency for group B is 45/500 = 0.09.

$$
\begin{array}{|c|c|c|} \hline
\mathrm{ABO} & & \mathrm{Relative} \\
\mathrm{Group} & \mathrm{Frequency} & \mathrm{Frequency} \\ \hline
\mathrm{A} & 210 & 0.42 \\  \hline
\mathrm{B} & 45  & 0.09 \\  \hline
\mathrm{AB} & 15 & 0.03 \\  \hline
\mathrm{O} & 230 & 0.46 \\  \hline
\end{array}\\
\mbox{Table 1: Frequency Table for ABO Group Data}
$$

## Pie Charts

The pie chart in Figure \@ref(fig:pie-chart) depicts the ABO group data. In a pie chart, 
each category is represented by a slice of the pie. The area of the slice is 
proportional to the percentage of items in the category -- that is, the 
relative frequency multiplied by 100.

```{r pie-chart, fig.cap="Relative Frequencies for ABO Blood Groups", warning=FALSE, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

data <- data.frame(ABO_Group = c("A", "B", "AB", "O"), value = c(210, 45, 15, 230)) %>%
  mutate(prop = value / sum(value) * 100)

ggplot(data = data, aes(x = "", y = prop, fill = ABO_Group)) + 
  geom_bar(stat = "identity", width = 1, color = "white") + 
  coord_polar("y", start = 0) + 
  theme_void()
```

Pie charts are effective for displaying the relative frequencies of a small 
number of categories. They are not recommended, however, when you have a large 
number of categories. Pie charts can also be confusing when they are used to 
compare the outcomes of two different surveys or experiments. In an influential 
book on the use of graphs, Edward Tufte asserted, "The only worse design than 
a pie chart is several of them."


## Bar Charts

Bar charts can also be used to represent frequencies of different categories. 
A bar chart of the ABO frequencies is shown in Figure \@ref(fig:bar-chart). 
Frequencies are shown 
on the Y-axis and the blood group is shown on the X-axis. 

```{r bar-chart, fig.cap="Frequencies for ABO Blood Groups", echo=FALSE}

ggplot(data, aes(ABO_Group, value)) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) + 
  labs(x = "ABO Group", y = "Frequency")
```

The Y-axis also can show the percentage of observations 
instead of the number of observations, as in Figure \@ref(fig:bar-chart-percent).

```{r bar-chart-percent, echo=FALSE, fig.cap="Percentages for ABO Blood Groups"}
ggplot(data, aes(ABO_Group, prop)) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) + 
  labs(x = "ABO Group", y = "Percentage")
```

## Comparing Distributions

Often we need to compare different sets of data, or different subsets
within the same overall data set In this case, we are comparing the 
"distributions" of outcomes or responses. Bar charts are 
often excellent for illustrating differences between two distributions.  Table
2 shows the distribution (in percentages) of ABO blood groups for those in
Albania and Australia.

$$
\begin{array}{|c|c|c|} \hline
\text{ABO} && \\
\text{Group} & \text{Albania} & \text{Australia} \\ \hline
\mathrm{A} & 36.7 & 38.0 \\  \hline
\mathrm{B} & 17.1 & 10.0 \\  \hline
\mathrm{AB} & 6.1 & 3.0 \\  \hline
\mathrm{O} & 40.1 & 49.0 \\  \hline
\end{array}\\
\mbox{Table 2: ABO Blood Group Percentages}
$$
From Table 2 we see that ABO groups B and AB are more common in Albania, 
group O is more common in Australia, and group A is similar for both.
This can be seen in the bar chart in Figure \@ref(fig:horiz-bar).

```{r horiz-bar, echo=FALSE, fig.cap="Percentages for ABO Blood Groups by Country"}
library(tidyr)

data <- data.frame(ABO_Group = c("A", "B", "AB", "O"), 
                   Albania = c(36.7, 17.1, 6.1, 40.1), 
                   Australia = c(38, 10, 3, 49)) %>%
  pivot_longer(!ABO_Group, names_to = "Country", values_to = "Percentage")

ggplot(data, aes(ABO_Group, Percentage, fill = Country)) + 
  geom_bar(stat = "identity", position = "dodge", width = 0.8) + 
  coord_flip() +
  labs(x = "ABO Group", y = "Percentage")
```

The bars in Figure \@ref(fig:horiz-bar) are oriented horizontally rather than vertically. The 
horizontal format is useful when you have many categories because there is more 
room for the category labels. 

---

## Exercises

1. $\text{Put exercises here}$


## Code Appendix

```{r, eval=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)

data <- data.frame(ABO_Group = c("A", "B", "AB", "O"), value = c(210, 45, 15, 230)) %>%
  mutate(prop = value / sum(value) * 100)

# Figure 7.1 -----
ggplot(data = data, aes(x = "", y = prop, fill = ABO_Group)) + 
  geom_bar(stat = "identity", width = 1, color = "white") + 
  coord_polar("y", start = 0) + 
  theme_void()

# Figure 7.2 -----
ggplot(data, aes(ABO_Group, value)) + 
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(x = "ABO Group", y = "Frequency")

# Figure 7.3 -----
ggplot(data, aes(ABO_Group, prop)) + 
  geom_bar(stat = "identity", fill = "#B4464B") + 
  labs(x = "ABO Group", y = "Percentage")

# Figure 7.4 -----
data <- data.frame(ABO_Group = c("A", "B", "AB", "O"), 
                   Albania = c(36.7, 17.1, 6.1, 40.1), 
                   Australia = c(38, 10, 3, 49)) %>%
  pivot_longer(!ABO_Group, names_to = "Country", values_to = "Percentage")

ggplot(data, aes(ABO_Group, Percentage, fill = Country)) + 
  geom_bar(stat = "identity", position = "dodge", width = 0.8) + 
  labs(x = "ABO Group")
  coord_flip()
```
