# Graphing Distributions: Histograms

 <font size="2"> Note: Portions below draw on content from
 *Online Statistics Education: A Multimedia Course of Study* 
 (http://onlinestatbook.com/) Project Leader: David M. Lane, Rice University </font>
 
## Introduction

The "distribution" of a set of data consists of the set of possible data values and the
frequency that the values occur.  Understanding how data is distributed is a
critical component of statistics and data analysis.  Visualizing data is an
effective way to better understand the distribution and there are a variety 
of visualization methods available.

A **histogram** is one graphical method for displaying the shape of a distribution
of quantitative data.
It is useful when there are a large number of data values, providing information
about the number of values in different groupings.  Histograms are widely used
and are fairly easy to produce with software.

## Frequency Tables

We begin with an example consisting of the weights of 371 bags containing a
substance suspected of being narcotics.  The weights of the bags range from
215 to 270 grams.  Below are the weights for 10 random bags.  (There are too many
values for it to be practical or useful to show them all.)

```{r, echo=FALSE}
set.seed(4)
data01 <- rnorm(371, mean = 239, sd = 8)
data01 <- data.frame(Weights = ifelse(data01 > 239, 239 + 1.1 * (data01 - 239), data01))
round(data01, 2)[1:10, 1]
```

One way to get a sense of the distribution of the weights is through the
use of a **frequency table,** where we organize the data into groups of similar
weights and give the count the number in each group.  The results for this
data set are shown in Table 9.1.

```{r, echo=FALSE}
table <- data.frame(a = seq(215, 260, by = 5), b = seq(220, 265, by = 5), f = c(4, 8, 31, 74, 88, 75, 57, 25, 8, 1))
table %>% 
  kbl(caption = "Grouped Frequency Distribution of Weights", 
      col.names = c("Interval Lower Limit", "Interval Upper Limit", "Class Frequency")) %>% 
  kable_styling(full_width = FALSE)
```

To create this table, the range of weights was broken into *class intervals*. 
The first interval is from 215 to 220, the second from 220 to 225, and so on. 
Next the number of weights
falling into each interval was counted to obtain the class frequencies. There are four weights  
in the first interval, eight in the second, and so on.

Class intervals of width 5 provide enough detail about the distribution without 
making the graph too "choppy." How one goes about choosing the widths (called *bin widths*) of 
class intervals is discussed later in this section. 

## Histograms

In a histogram the class frequencies are represented by bars with 
the height of each bar corresponds to its class frequency,
providing a graphical
depiction of the data.  The height of each bar corresponds 
to its class frequency. A histogram of the weight data is shown in 
Figure \@ref(fig:histogram).

```{r histogram, fig.cap="Histogram of Weights", echo=FALSE}
data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights), breaks = seq(215, 265, by = 5), fill = "steelblue", color = "black") + 
  labs(y = "Frequency")
```

We can see from the histogram that most of the weights are between 230 and 250 grams,
with fewer weights in the extremes.  We can also see that the distribution is not 
quite symmetric, with the weights extending farther to the right than to the left.  This
distribution can be said to be *skewed*. 

Because the weights are measured to two decimal 
places, it is not likely that we get many "fence-sitters" -- that is, values
that land right on the border between two bins.  This allows us to choose
whole numbers as boundaries,  avoiding a cluttered appearance.  Computer
software generally will use this option when feasible, and sometimes automatically
labels the middle of each interval instead of the endpoints.
(Software typically allows for the user to specify bins if desired.)

Histograms can be based on relative frequencies instead of actual frequencies,
showing the proportion (or percentage) of values in each interval instead of
the exact frequency.  Our weight data is displayed as percentages in 
Figure \@ref(fig:histogram-perc).  To get the percentages, we divide
each frequency by the number of data values (that gives the proportion) and
then multiply by 100 to convert to percentages.

```{r histogram-perc, fig.cap="Histogram of Weights", echo=FALSE}
data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights, y = 100 * ..count.. / sum(count)), 
                 breaks = seq(215, 265, by = 5), 
                 fill = "steelblue", color = "black") + 
  labs(y = "Percentage")
```

We see that the shape of the histogram is unchanged, only the Y-axis is
different.

## Bin Widths

The choice of bin width determines the number of class intervals. 
This decision, along with the choice of starting point for the first interval, 
affects the shape of the histogram. In Figure \@ref(fig:histogram-two) we see
histograms for our data using bins of width 1 (left) and width 10 (right).
These both give a general sense of the distribution, but here bins of width 1
produce a graph that is too busy while bins of width 10 produce a graph that
lacks detail.

```{r histogram-two, fig.cap="Histograms with Different Bin Widths", echo=FALSE, fig.show='hold', out.width="50%"}
data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights), binwidth = 1, fill = "steelblue", color = "black") + 
  labs(y = "Frequency")

data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights), binwidth = 10, fill = "steelblue", color = "black") + 
  labs(y = "Frequency")
```

Most software has defaults for bin widths that might work well for a given set
of data, or might not.  There are some "rules of thumb" that can help guide
the choice of bin width but these are just guidelines that should not be
regarded as definitive, so do not feel bound to them.  That said, here are
a couple of general guidelines you can consider:

$\bullet$ Sturges' rule is to set the number of intervals as close as possible to 
$1 + \log_2(N)$, where $\log_2(N)$ is the base 2 log of the number of 
observations $N$. For our data set of $N = 371$, we have $1 + \log_2(N) \approx 9.535$,
suggesting 10 intervals which happens to match our choice.

$\bullet$ The Rice rule suggests that the number of intervals should be the integer
nearest to $2\sqrt[3]{N},$ twice the cube root of the number of data values.
For $N = 371$ this formula gives $2\sqrt[3]{371} \approx 14.371$, which is somewhat more
than we chose.

The above rules are just a guide. You are advised to experiment with different 
numbers of intervals and choose the histogram that you feel best conveys the
shape of the distribution.

---

## Exercises

1. $\text{Put exercises here}$

## Code Appendix

```{r, eval=FALSE}
set.seed(4)
data01 <- rnorm(371, mean = 239, sd = 8)
data01 <- data.frame(Weights = ifelse(data01 > 239, 239 + 1.1 * (data01 - 239), data01))

# Figure 8.1 -----
data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights), breaks = seq(215, 265, by = 5), 
                 fill = "steelblue", color = "black") + 
  labs(y = "Frequency")

# Figure 8.2 -----
data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights, y = 100 * ..count.. / sum(count)), 
                 breaks = seq(215, 265, by = 5), 
                 fill = "steelblue", color = "black") + 
  labs(y = "Percentage")

# Figure 8.3 -----
data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights), binwidth = 1, fill = "steelblue", color = "black") + 
  labs(y = "Frequency")

data01 %>% ggplot() + 
  geom_histogram(aes(x = Weights), binwidth = 10, fill = "steelblue", color = "black") + 
  labs(y = "Frequency")
```

