# Graphing Distributions: Box Plots

 <font size="2"> Note: Portions below modeled after content from
 *Online Statistics Education: A Multimedia Course of Study* 
 (http://onlinestatbook.com/) Project Leader: David M. Lane, Rice University </font>
 
## Basic box plots

In this chapter we discuss the *box plot* which provides a useful way to graphically
display information about the distribution of a set of data,
identify outliers, and compare distributions.

To get us started, suppose that we have small pieces of glass from two different
sources: a broken window at the scene of a burglary and the trunk of a car
belonging to a suspect.  To compare the glass from the two sources, a trace 
element present in glass can be measured.  The amount present varies within
a sheet of glass so measurements are taken from numerous pieces found at the
crime scene and in the car trunk.  (The specifics of the element and measurement
units are not important for this discussion.)  The measurements are given below:

Car Trunk:
```{r, echo=FALSE}
data02 <- c(67.9, 68.9, 69.5, 67.6, 73.4, 64.5, 73.0, 59.7, 66.9, 68.2, 71.8, 64.8, 56.3)
data02
```

Crime Scene:
```{r, echo=FALSE}
data01 <- c(55.6, 65.6, 68.0, 68.9, 67.7, 66.5, 66.8, 60.3, 68.0, 71.6, 61.4, 63.2, 68.2, 64.1, 64.8)
data01
```

To construct the box plot, we start by finding the 25th, 50th, and 75th percentiles
for each of our data sets.  Recall that the 25th percentile is the quantity such
that 25% of the data values are less than this quantity, and similarly for the
other percentiles.  The percentiles for each of our data sets are shown in
Table 10.1.

```{r, echo=FALSE}
table <- data.frame(a = c("$25^{th}$", "$50^{th}$", "$75^{th}$"), b = c(64.80, 67.90, 69.50), d = c(63.65, 66.50, 68.00))
table %>% kbl(caption = "Percentiles", col.names = c("Percentile", "Car Trunk", "Crime Scene")) %>% 
  kable_styling(full_width = FALSE)
```

Figure \@ref(fig:boxplot1) shows how these percentiles 
are incorporated into box plots.  The lower and upper 
limits of the box (called "hinges") extend from the 
25th to 75th percentile and a line between those at the 50th percentile 
(which is the median).

```{r boxplot1, fig.cap="Box Plots of Glass Measurements", echo=FALSE}
## Joseph: The legend runs off the side of the page in the HTML output.
##         Can you see if you can figure out a way to fix that?

data01 <- data.frame(Measurement = data01, Source = "Crime Scene")
data02 <- data.frame(Measurement = data02, Source = "Car Trunk")
data03 <- rbind(data01, data02)

data03 %>% ggplot(aes(x = Source, y = Measurement)) +
  geom_boxplot(outlier.shape = NA, coef = 0)
```

Comparing box plots, we see that the percentiles for the car trunk data
are somewhat greater than for the crime scene data but there is considerable
overlap between them.  Based on this, it is difficult to say whether the 
glass pieces from car trunk and crime scene are from the same
original source or different sources.

Most box plots have more information than that shown in Figure \@ref(fig:boxplot1)
and there are a wide variety of options that can be included.  We give a few
examples below to provide a sense of what is possible but these are by no means
comprehensive.

We next add "whiskers" above and below each box to give additional information 
about the spread of the data. Whiskers are vertical lines that (optionally)
end in a horizontal stroke. Define the "step" to be 1.5 times the
IQR = 75th percentile - 25th percentile (IQR = "Interquartile range"). 
The upper whisker extends
from the upper hinge to the greatest data value that is less than one "step"
above the upper hinge, with the lower whisker similarly defined.  (Software
automates the process so one need not do this by hand.)

Figure \@ref(fig:boxplot2) shows the box plots with whiskers.

```{r boxplot2, fig.cap="Box Plots of Glass Measurements", echo=FALSE}
## Joseph: The legend runs off the side of the page in the HTML output.
##         Can you see if you can figure out a way to fix that?

data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot(outlier.shape = NA)
```

The whiskers add additional information about the spread of data, but do
not extend to all values.  We indicate these with dots that are beyond 
the whiskers in positions corresponding to data values.  This is shown in
Figure \@ref(fig:boxplot3).

```{r boxplot3, fig.cap="Box Plots of Glass Measurements", echo=FALSE}
## Joseph: The legend runs off the side of the page in the HTML output.
##         Can you see if you can figure out a way to fix that?

data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot(outlier.size = 2)
```

Some (but not all!) box plots also include a mark indicating the location of
the mean.  This is added in Figure \@ref(fig:boxplot4) as a red dot in each
box plot.

```{r boxplot4, fig.cap="Box Plots of Glass Measurements", echo=FALSE}
## Joseph: The legend runs off the side of the page in the HTML output.
##         Can you see if you can figure out a way to fix that?

data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") +
  geom_boxplot(outlier.size = 2) + 
  stat_summary(fun = mean, geom = "point", color = "red", size = 2)
```

Figure \@ref(fig:boxplot4) provides a good summary of the data. Half of 
the scores in a distribution are between the hinges (recall that the hinges are 
the 25th and 75th percentiles), we see that half the measurements for the glass
from the car trunk are between about 65 and 69, while the measurements from the
crime scene are between about 63.5 and 68.  The IQR for each is about the same
but the overall range for the car trunk measurements is somewhat greater 
then the crime scene measurements.  For both data sets the mean is less than
the median, which is to be expected due to the extreme small values displayed
in Figure \@ref(fig:boxplot4).

## Variations on box plots

Statistical analysis programs typically offer numerous options for creating 
box plots. For instance, Figure \@ref(fig:boxplot5) shows the box plots 
as above but also includes the individual measurements along the vertical
axis.  (The means are still red to distinguish it from the data values.)
This provides more detail on our data -- note the points that specify 
the ends of the whiskers as well as the difference in how the points are
distributed within the boxes.

```{r boxplot5, fig.cap="Box Plots of Glass Measurements", echo=FALSE}
## Joseph: The legend runs off the side of the page in the HTML output.
##         Can you see if you can figure out a way to fix that?

data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot() + geom_jitter(width = 0, alpha = 0.9) + 
  stat_summary(fun = mean, geom = "point", color = "red", size = 2)
```

Showing the specific data values on the vertical axis works well as long as there 
are not too many points and they do not tend to stack on top of each other.
In the case of more points and more repetition of values, an option that can
work is "jittering" which adds a horizontal offset so that points can be
distinguished. Figure \@ref(fig:boxplot6) shows our data with a modest 
amount of jitter added.

```{r boxplot6, fig.cap="Box Plots of Glass Measurements", echo=FALSE}
## Joseph: The legend runs off the side of the page in the HTML output.
##         Can you see if you can figure out a way to fix that?

data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot() + geom_jitter(width = 0.1, alpha = 0.9) + 
  stat_summary(fun = mean, geom = "point", color = "red", size = 2)
```

Different styles of box plots are best for different situations, and there are 
no firm rules for which to use. When exploring your data, you should try several 
ways of visualizing them. Which graphs you include in your report should depend 
on how well different graphs reveal the aspects of the data you consider most 
important.

---

## Exercises

1. $\text{Put exercises here}$

## Code Appendix

```{r, eval=FALSE}
data01 <- c(55.6, 65.6, 68.0, 68.9, 67.7, 66.5, 66.8, 60.3, 68.0, 71.6, 61.4, 63.2, 68.2, 64.1, 64.8)
data02 <- c(67.9, 68.9, 69.5, 67.6, 73.4, 64.5, 73.0, 59.7, 66.9, 68.2, 71.8, 64.8, 56.3)

data01 <- data.frame(Measurement = data01, Source = "Crime Scene")
data02 <- data.frame(Measurement = data02, Source = "Car Trunk")
data03 <- rbind(data01, data02)

# Figure 9.1 -----
data03 %>% ggplot(aes(x = Source, y = Measurement)) +
  geom_boxplot(outlier.shape = NA, coef = 0)

# Figure 9.2 -----
data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot(outlier.shape = NA)

# Figure 9.3 -----
data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot()

# Figure 9.4 -----
data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") +
  geom_boxplot(outlier.size = 2) + 
  stat_summary(fun = mean, geom = "point", color = "red", size = 2)

# Figure 9.5 -----
data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot() + geom_jitter(width = 0, alpha = 0.9) + 
  stat_summary(fun = mean, geom = "point", color = "red", size = 2)

# Figure 9.6 -----
data03 %>% ggplot(aes(x = Source, y = Measurement)) + stat_boxplot(geom = "errorbar") + 
  geom_boxplot() + geom_jitter(width = 0.1, alpha = 0.9) + 
  stat_summary(fun = mean, geom = "point", color = "red", size = 2)
```
