# Graphing Distributions: Bar Charts

 <font size="2"> Note: Portions below modeled after content from
 *Online Statistics Education: A Multimedia Course of Study* 
 (http://onlinestatbook.com/) Project Leader: David M. Lane, Rice University </font>


In the section on qualitative variables, we saw how bar charts can be used 
to illustrate the frequencies (or relative frequencies) of different categories. 
For example, the bar chart shown in Figure \@ref(fig:bar-chart-percent2) 
depicts the percentage of people in a population that have each of the four
possible ABO blood groups.

```{r bar-chart-percent2, fig.cap="Percentages for ABO Blood Groups", echo=FALSE}

library(dplyr)
library(ggplot2)

data <- data.frame(ABO_Group = c("A", "B", "AB", "O"), value = c(210, 45, 15, 230)) %>%
  mutate(prop = value / sum(value) * 100)

ggplot(data, aes(ABO_Group, prop)) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) + 
  labs(x = "ABO Group", y = "Percentage")
```

In this chapter we consider how bar charts can be used to present other
types of quantitative information.  The bar chart in Figure \@ref(fig:bar-chart2)
presents the number of murders in four large U.S. cities. 

```{r bar-chart2, fig.cap="Number of murders, selected U.S. cities, year 2000", warning=FALSE, echo=FALSE, message=FALSE}

data <- data.frame(City = c("Chicago", "Dallas", "Los Angeles", "New York"), 
                   Murders = c(774, 251, 349, 468))

ggplot(data, aes(City, Murders)) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) + 
  labs(x = "City", y = "Murders")
```

A case can be made that more relevant than the total number of murders is
the per capita number of murders.  Figure \@ref(fig:bar-chart3) gives the
number of murders per 1,000,000 population for the same cities.

```{r bar-chart3, fig.cap="Murders per 1,000,000 residents, selected U.S. cities, year 2000", warning=FALSE, echo=FALSE, message=FALSE}

data <- data.frame(City = c("Chicago", "Dallas", "Los Angeles", "New York"), 
                   Murders = c(281.8645, 192.4847, 89.51013, 53.15766)) 

ggplot(data, aes(City, Murders)) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) + 
  labs(x = "City", y = "Murders per 1,000,000 residents")
```

Bar charts can be useful for showing change over time. Figure \@ref(fig:bar-chart4), 
shows the year-to-year percent change in the number of shootings in the city of
Toronto. The considerable variation from one year to the next is clear.  Note also
that negative percentages are possible, the result of the number of shootings 
decreasing from one year to the next.

```{r bar-chart4, fig.cap="Year-to-year Percent Change, Shootings in Toronto", warning=FALSE, echo=FALSE, message=FALSE}

data <- data.frame(City = c("1995", "1996", "1997", "1998", "1999", "2000"), 
                   RateChange = c(62.7, 41.3, -2.9, 7.3, 16.7, -6.7)) 

# 2014-2020: 177, 288, 407, 395, 424, 495, 462 Shootings in Toronto (Wikipedia)

ggplot(data, aes(City, RateChange)) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) + 
  labs(x = "Year", y = "Year-to-year Percent Change, Shootings")
```

Bar charts are often used to compare the means of different experimental 
conditions. Figure \@ref(fig:bar-chart5) shows the mean trace element 
measurement for collections of glass pieces found at a crime scene and
in the car trunk of a suspect.  (This data is treated more thoroughly
in the chapter on box plots.)

```{r bar-chart5, fig.cap="Comparison of mean trace element measurements", warning=FALSE, echo=FALSE, message=FALSE}

data01 = c(55.6, 65.6, 68.0, 68.9, 67.7, 66.5, 66.8, 60.3, 68.0, 71.6, 61.4, 63.2, 68.2, 64.1, 64.8)
data02 = c(67.9, 68.9, 69.5, 67.6, 73.4, 64.5, 73.0, 59.7, 66.9, 68.2, 71.8, 64.8, 56.3)
data <- data.frame(Source = c("Car Trunk", "Crime Scene"), 
                   Mean = c(mean(data02), mean(data01)))

ggplot(data, aes(Source, Mean)) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) + 
  labs(x = "Source", y = "Mean trace element measurement")
```

The heights of the bars are cery similar, suggesting that there might be 
a common source for the two sets of glass pieces.  That might well be the
case but the means alone are not enough to provide a conclusion one way or
the other.  In this case, box plots can be used instead since they provide 
more information about the distribution.  Figure \@ref(fig:boxplot2a)
shows the box plots for this data again.

```{r boxplot2a, fig.cap="Box plots of glass measurements", echo=FALSE}

## Joseph: The legend runs off the side of the page in the HTML output.
##         Can you see if you can figure out a way to fix that?
## Joseph: Can you add the horizontal line to the end of the whiskers?

data01 = c(55.6, 65.6, 68.0, 68.9, 67.7, 66.5, 66.8, 60.3, 68.0, 71.6, 61.4, 63.2, 68.2, 64.1, 64.8)
data02 = c(67.9, 68.9, 69.5, 67.6, 73.4, 64.5, 73.0, 59.7, 66.9, 68.2, 71.8, 64.8, 56.3)

data01 <- data.frame(Measurement = data01)
data01['Source'] = 'Crime Scene'

data02 <- data.frame(Measurement = data02)
data02['Source'] = 'Car Trunk'

data03 = rbind(data01, data02)

ggplot(data03, aes(x=Source, y=Measurement, color=Source)) + 
  geom_boxplot(outlier.shape = NA)

```

The box plots have considerable overlap, but at the same time the middle 50%
of values from the car trunk seem to trend higher than those from the crime
scene.  This suggests the possibility of different sources for the two sets
of glass pieces while not ruling out the possibilty of them coming from the
same source.  We will consider this question more in later chapters.

---

## Exercises

1. $\text{Put exercises here}$













